A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol

### 2024-01-07 paper.tex
reduce to 32 + 3 references (almost there)

\textcolor{red}{Scatterplots of residuals against the fitted values, and each of the explanatory variables, are commonly used to scrutinize their relationships}.
If there are any visually discoverable associations, the model is
potentially inadequate or incorrectly specified.
\textcolor{red}{We can also potentially discover patterns not directly connected to a linear model assumption from these residual plots, such as the discreteness or skewness of the fitted values, and outliers.}
To read residual plots, one looks for noticeable departures from the
model such as non-linear pattern or heteroskedasticity. A non-linear
pattern would suggest that the model needs to have some additional
non-linear terms. Heteroskedasticity suggests that the error is
dependent on the predictors, and hence violates the independence
assumption. Statistical tests were developed to provide objective
assessment, for example, of non-linear patterns
\citep[e.g.][]{ramsey1969tests}, and heteroskedasticity
\citep[e.g.][]{breusch1979simple}.

@@

During the experiment, every participant recruited from the Prolific
crowd-sourcing platform \citep{palan2018prolific} is presented with a
block of 20 lineups. A lineup consists of a randomly placed data plot
and 19 null plots, which are all residual plots drawn with raw residuals
on the y-axis and fitted values on the x-axis. An additional horizontal
red line is added at \(y = 0\) as a visual reference. The data in the
data plot is simulated from one of two models described in Section
\ref{simulating-departures-from-good-residuals}, while the data of the
remaining 19 null plots are generated by the residual rotation technique
discussed in Section \ref{visual-test-procedure-based-on-lineups}.

@@

Overall, we collected 7974 evaluations on 1152 unique lineups performed
by 443 participants throughout three data collection periods. Details
about lineup allocation and data collection are provided in the
Appendix.

A summary of the factors used in the experiment can be found in Table
\ref{tab:model-factor-table}. There were four levels of the non-linear
structure, and three levels of heteroskedastic structure. The signal
strength was controlled by error variance (\(\sigma\)) for the
non-linear pattern, and by a ratio (\(b\)) parameter for the
heteroskedasticity. Additionally, three levels of sample size (\(n\))
and four different fitted value distributions were incorporated.

@@

Data collection period I was designed to study the ability of
participants to detect the effect of a non-linear term
\(\boldsymbol{z}\) constructed using Hermite polynomials on random
vector \(\boldsymbol{x}\) formulated as

\[ \boldsymbol{y} = \boldsymbol{1}_n + \boldsymbol{x} + \boldsymbol{z} + \boldsymbol{\varepsilon},~\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \sigma^2\boldsymbol{I}_n)\]

\noindent where \(\boldsymbol{y}\), \(\boldsymbol{x}\),
\(\boldsymbol{\varepsilon}\), \(\boldsymbol{z}\) are vectors of size
\(n\), \(\boldsymbol{1}_n\) is a vector of ones of size \(n\),
\(\boldsymbol{z} \propto He_j(\boldsymbol{x})\), \(He_{j}(.)\) is the
\(j\)th-order probabilist's Hermite polynomials
\citetext{\citealp{hermite1864nouveau}; \citealp[originally
by][]{de1820theorie}}.

To fit a misspecified model, the \(\boldsymbol{z}\) term, which is a
higher order term of \(\boldsymbol{x}\), was omitted from the null
regression formula

\begin{equation} \label{eq:null-model}
\boldsymbol{y} = \beta_0\boldsymbol{1}_n + \beta_1 \boldsymbol{x} + \boldsymbol{u}, \boldsymbol{u} \sim N(\boldsymbol{0}_n, \sigma^2\boldsymbol{I}_n).
\end{equation}

Four different values of \(j = 2, 3, 6, 18\) were chosen so that
distinct shapes of non-linearity were included in the residual plot.
These include ``U'', ``S'', ``M'' and ``triple-U'' shape as shown in
Figure \ref{fig:different-shape-of-herimite}. A greater value of \(j\)
will result in a curve with more turning points. It is expected that the
``U'' shape will be the easiest to detect, and as the shape gets more
complex it will be harder to perceive in a scatterplot, particularly
when there is noise. Figure \ref{fig:different-sigma} shows the ``U''
shape for different amounts of noise (\(\sigma\)).

Data collection period II was similar to Data collection I but focuses
on heteroskedasticity visual patterns formulated as

\[\boldsymbol{y} = \boldsymbol{1}_n + \boldsymbol{x} + \boldsymbol{\varepsilon},~\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \boldsymbol{1}_n + (2 - |a|)(\boldsymbol{x} - a\boldsymbol{1}_n)'(\boldsymbol{x} - a\boldsymbol{1}_n) b \boldsymbol{I}_n),\]
\noindent where \(\boldsymbol{y}\), \(\boldsymbol{x}\),
\(\boldsymbol{\varepsilon}\) are vectors of size \(n\).

The null regression model used to fit the realizations generated by the
above model is formulated exactly the same as Equation
\ref{eq:null-model}. For \(b \neq 0\), the variance-covariance matrix of
the error term \(\boldsymbol{\varepsilon}\) is correlated with the
predictor \(\boldsymbol{x}\), which will lead to the presence of
heteroskedasticity.

Visual patterns of heteroskedasticity are simulated using three
different shapes (\(a\) = -1, 0, 1) including ``left-triangle'',
``butterfly'' and ``right-triangle'' shapes as displayed in Figure
\ref{fig:different-shape-of-heter}. Figure \ref{fig:different-b} shows
the butterfly shape as the ratio parameter (\(b\)) is changed.

More details about the design of the data generating process are
provided in the Appendix.

@@

Fitted values are a function of the independent variables, and the
distribution of the observed values affects the distribution of the
fitted values. In the best case scenario the fitted values will have a
uniform distribution, which means that there is even coverage of
possible observed values across all of the predictors. This is not
always present in the collected data. Sometimes the fitted values are
discrete because one or more predictors were measured discretely. It is
also common to see a skewed distribution of fitted values if one or more
of the predictors has a skewed distribution. This latter problem is
usually corrected before modelling, using a variable transformation. Our
simulation assess this by using four different distributions to
represent fitted values, constructed by different sampling of the raw
predictor \(X_{raw}\), namely, uniform (\(U(-1, 1)\)), normal
(\(N(0, 0.3^2)\)), skewed (\(\text{lognormal}(0, 0.6^2)/3\)) and
discrete (\(U\{1, 5\}\)).

@@

Three different sample sizes were used in our experiment:
\(n = 50, 100, 300\). Figure \ref{fig:different-n} shows the non-linear
``S'' shape for different sample sizes. We expect signal strength to
decline in the simulated data plots with smaller \(n\). We chose 300 as
the upper limit, because it is typically enough for structure to be
visible in a scatterplot reliably.
\textcolor{red}{Beyond 300, the scatterplot should probably be used with transparency or replaced with a density or binned plot as scatterplots suffer from over-plotting.}

@@

For the non-linearity model, the key items defining effect size are
sample size (\(n\)) and variance of the error term (\(\sigma^2\)), and
so effect size would be roughly calculated as \(\sqrt{n}/{\sigma}\). As
sample size increases the effect size would increase, but as variance
increases the effect size decreases. However, it is not clear how the
additional parameter for the model polynomial order, \(j\), should be
incorporated. Intuitively, the large \(j\) means more complex pattern,
which likely means effect size would decrease. For the purposes of our
calculations we have chosen to use an approach based on Kullback-Leibler
divergence \citep{kullback1951information}, coupled with simulation.
This formulation defines effect size to be:

\begin{equation}
E = \frac{1}{2}\left(\boldsymbol{\mu}_z'(\text{diag}(\boldsymbol{R}\sigma^2))^{-1}\boldsymbol{\mu}_z\right)
\label{eq:nonlin-effect-size}
\end{equation}

\noindent where \(\text{diag}(.)\) is the diagonal matrix constructed
from the diagonal elements of a matrix,
\(\boldsymbol{R} = \boldsymbol{I}_n - \boldsymbol{H}\) is the residual
operator,
\(\boldsymbol{H} = \boldsymbol{X}(\boldsymbol{X}'\boldsymbol{X})^{-1}\boldsymbol{X}'\)
is the hat matrix,
\(\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z\)
is the expected values of residuals where \(\boldsymbol{Z}\) contain any
higher order terms of \(\boldsymbol{X}\) left out of the regression
equation, \(\boldsymbol{\beta}_z\) contains the corresponding
coefficients, and \(\sigma^2\boldsymbol{I}_n\) is the assumed covariance
matrix of the error term when \(H_0\) is true.

In the heteroskedasticity model, the key elements for measuring effect
size are sample size, \(n\), and the ratio of the biggest variance to
smallest variance, \(b\). Larger values of both would produce higher
effect size. However, it is not clear how to incorporate the additional
shape parameter, \(a\). Thus the same approach is used here, where the
formula can be written as:

\begin{equation}
E = \frac{1}{2}\left(\log\frac{|\text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}')|}{|\text{diag}(\boldsymbol{R})|} - n + \text{tr}(\text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}')^{-1}\text{diag}(\boldsymbol{R}))\right)
\label{eq:hetero-effect-size}
\end{equation}

\noindent where \(\boldsymbol{V}\) is the actual covariance matrix of
the error term. Derivations for these equations are provided in the
Appendix.

@@

\hypertarget{section}{%
\section{\texorpdfstring{\textcolor{red}{Limitations and practicality}}{}}\label{section}}

@@

These \texttt{R} packages are used for the work: \texttt{tidyverse}
\citep{tidyverse}, \texttt{lmtest} \citep{lmtest}, \texttt{mpoly}
\citep{mpoly}, \texttt{ggmosaic} \citep{ggmosaic}, \texttt{kableExtra}
\citep{kableextra}, \texttt{patchwork} \citep{patchwork},
\texttt{rcartocolor} \citep{rcartocolor}. The study website is powered
by \texttt{PythonAnywhere} \citep{pythonanywhere} and the \texttt{Flask}
web framework \citep{flask}. The \texttt{jsPsych} framework
\citep{jspsych} is used to create behavioural experiments that run in
our study website.

>>>>

Scatterplots of residuals against the fitted values, and each of the
explanatory variables, are commonly used to scrutinize their
relationships. If there are any visually discoverable associations, the
model is potentially inadequate or incorrectly specified. We can also
potentially discover patterns not directly connected to a linear model
assumption from these residual plots, such as the discreteness or
skewness of the fitted values, and outliers. To read residual plots, one
looks for noticeable departures from the model such as non-linear
pattern or heteroskedasticity. A non-linear pattern would suggest that
the model needs to have some additional non-linear terms.
Heteroskedasticity suggests that the error is dependent on the
predictors, and hence violates the independence assumption. Statistical
tests were developed to provide objective assessment, for example, of
non-linear patterns \citep[e.g.][]{ramsey1969tests}, and
heteroskedasticity \citep[e.g.][]{breusch1979simple}.

@@

During the experiment, every participant recruited from the Prolific
crowd-sourcing platform \citep{palan2018prolific} was presented with a
block of 20 lineups. A lineup consisted of a randomly placed data plot
and 19 null plots, which were all residual plots drawn with raw
residuals on the y-axis and fitted values on the x-axis. An additional
horizontal red line was added at \(y = 0\) as a visual reference. The
data in the data plot was simulated from one of two models described in
Section \ref{simulating-departures-from-good-residuals}, while the data
of the remaining 19 null plots were generated by the residual rotation
technique discussed in Section
\ref{visual-test-procedure-based-on-lineups}.

@@

Overall, we collected 7974 evaluations on 1152 unique lineups performed
by 443 participants. A summary of the factors used in the experiment can
be found in Table \ref{tab:model-factor-table}. There were four levels
of the non-linear structure, and three levels of heteroskedastic
structure. The signal strength was controlled by error variance
(\(\sigma\)) for the non-linear pattern, and by a ratio (\(b\))
parameter for the heteroskedasticity. Additionally, three levels of
sample size (\(n\)) and four different fitted value distributions were
incorporated.

@@

Data collection period I was designed to study the ability of
participants to detect non-linearity departures from residual plots. The
non-linearity departure was constructed by omitting a \(j\)th order
Hermite polynomial
\citetext{\citealp{hermite1864nouveau}; \citealp[originally
by][]{de1820theorie}} term of the predictor from the simple linear
regression equation. Four different values of \(j = 2, 3, 6, 18\) were
chosen so that distinct shapes of non-linearity were included in the
residual plots. These include ``U'', ``S'', ``M'' and ``triple-U'' shape
as shown in Figure \ref{fig:different-shape-of-herimite}. A greater
value of \(j\) will result in a curve with more turning points. It is
expected that the ``U'' shape will be the easiest to detect, and as the
shape gets more complex it will be harder to perceive in a scatterplot,
particularly when there is noise. Figure \ref{fig:different-sigma} shows
the ``U'' shape for different amounts of noise (\(\sigma\)).

Data collection period II was similar to period I but focuses on
heteroskedasticity departures. We generated the heteroskedasticity
departures by setting the variance-covariance matrix of the error term
as a function of the predictor, but fitted the data with the simple
linear regression model, intentionally violated the constant variance
assumption. Visual patterns of heteroskedasticity are simulated using
three different shapes (\(a\) = -1, 0, 1) including ``left-triangle'',
``butterfly'' and ``right-triangle'' shapes as displayed in Figure
\ref{fig:different-shape-of-heter}. Figure \ref{fig:different-b} shows
the butterfly shape as the ratio parameter (\(b\)) is changed. More
details about the simulation process are provided in the Appendix.

@@

Fitted values are a function of the independent variables, and the
distribution of the observed values affects the distribution of the
fitted values. In the best case scenario the fitted values will have a
uniform distribution, which means that there is even coverage of
possible observed values across all of the predictors. This is not
always present in the collected data. Sometimes the fitted values are
discrete because one or more predictors were measured discretely. It is
also common to see a skewed distribution of fitted values if one or more
of the predictors has a skewed distribution. This latter problem is
usually corrected before modelling, using a variable transformation. Our
simulation assess this by using four different distributions to
represent fitted values, constructed by different sampling of the
predictor, including \(U(-1, 1)\) (uniform), \(N(0, 0.3^2)\) (normal),
\(\text{lognormal}(0, 0.6^2)/3\) (skewed) and \(U\{-1, 1\}\) (discrete).

@@

Three different sample sizes were used in our experiment:
\(n = 50, 100, 300\). Figure \ref{fig:different-n} shows the non-linear
``S'' shape for different sample sizes. We expect signal strength to
decline in the simulated data plots with smaller \(n\). We chose 300 as
the upper limit, because it is typically enough for structure to be
visible in a scatterplot reliably. Beyond 300, the scatterplot should
probably be used with transparency or replaced with a density or binned
plot as scatterplots suffer from over-plotting.

@@

For the non-linearity model, the key items defining effect size are
sample size (\(n\)) and the noise level (\(\sigma^2\)), and so effect
size would be roughly calculated as \(\sqrt{n}/{\sigma}\). Increasing
sample size tends to boost the effect size, while heightened noise
diminishes it. However, it is not clear how the additional parameter for
the model polynomial order, \(j\), should be incorporated. Intuitively,
the large \(j\) means more complex pattern, which likely means effect
size would decrease. Similarly, in the heteroskedasticity model, effect
size relies on sample size (\(n\)) and the ratio of the largest to
smallest variance, \(b\). Larger values of both would produce higher
effect size, but the role of the additional shape parameter, \(a\), in
this context is unclear.

For the purposes of our calculations we have chosen to use an approach
based on Kullback-Leibler divergence \citep{kullback1951information},
coupled with simulation. This formulation defines effect size to be

\vspace{-\baselineskip}

\[E = \frac{1}{2}\left(\log\frac{|\text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}')|}{|\text{diag}(\boldsymbol{R}\sigma^2)|} - n + \text{tr}(\text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}')^{-1}\text{diag}(\boldsymbol{R}\sigma^2)) + \boldsymbol{\mu}_z'(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}')^{-1}\boldsymbol{\mu}_z\right),\]

\noindent where \(\text{diag}(.)\) is the diagonal matrix constructed
from the diagonal elements of a matrix, \(\boldsymbol{X}\) is the design
matrix, \(\boldsymbol{V}\) is the actual covariance matrix of the error
term,
\(\boldsymbol{R} = \boldsymbol{I}_n - \boldsymbol{X}(\boldsymbol{X}'\boldsymbol{X})^{-1}\boldsymbol{X}'\)
is the residual operator,
\(\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z\)
is the expected values of residuals where \(\boldsymbol{Z}\) contains
any higher order terms of \(\boldsymbol{X}\) left out of the regression
equation, \(\boldsymbol{\beta}_z\) contains the corresponding
coefficients, and \(\sigma^2\boldsymbol{I}_n\) is the assumed covariance
matrix of the error term when \(H_0\) is true.

@@

\hypertarget{limitations-and-practicality}{%
\section{Limitations and
practicality}\label{limitations-and-practicality}}

@@

These \texttt{R} packages were used for the work: \texttt{tidyverse}
\citep{tidyverse}, \texttt{lmtest} \citep{lmtest}, \texttt{mpoly}
\citep{mpoly}, \texttt{ggmosaic} \citep{ggmosaic}, \texttt{kableExtra}
\citep{kableextra}, \texttt{patchwork} \citep{patchwork},
\texttt{rcartocolor} \citep{rcartocolor}. The study website was powered
by \texttt{PythonAnywhere} \citep{pythonanywhere} and \texttt{Flask} web
framework \citep{flask}. The \texttt{jsPsych} framework \citep{jspsych}
was used to create behavioural experiments that run in our study
website.
